EEG Waveform Analysis of P300 ERP with Applications to Brain Computer Interfaces

### 2018-11-27 brainsci-371413-done/brainsci-371413.tex
Final and published version

%=================================================================
\documentclass[brainsci,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}

@@

%
%--------------------
% Class Options:
%--------------------
% journal
%----------
% Choose between the following MDPI journals:
% actuators, admsci, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, atmosphere, atoms, axioms, batteries, behavsci, beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci, buildings, carbon, cancers, catalysts, cells, challenges, chemosensors, children, chromatography, climate, coatings, computation, computers, condensedmatter, cosmetics, cryptography, crystals, data, dentistry, designs, diagnostics, diseases, diversity, econometrics, economies, education, electronics, energies, entropy, environments, epigenomes, fermentation, fibers, fishes, fluids, foods, forests, futureinternet, galaxies, games, gels, genealogy, genes, geosciences, geriatrics, healthcare, horticulturae, humanities, hydrology, informatics, information, infrastructures, inorganics, insects, instruments, ijerph, ijfs, ijms, ijgi, ijtpp, inventions, jcdd, jcm, jdb, jfb, jfmk, jimaging, jof, jintelligence, jlpea, jmse, jpm, jrfm, jsan, land, languages, laws, life, literature, lubricants, machines, magnetochemistry, marinedrugs, materials, mathematics, mca, mti, medsci, medicines, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, molbank, molecules, mps, nanomaterials, ncrna, neonatalscreening, nutrients, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, plants, polymers, processes, proteomes, publications, recycling, religions, remotesensing, resources, risks, robotics, safety, sensors, separations, sexes, sinusitis, socsci, societies, soils, sports, standards, sustainability, symmetry, systems, technologies, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vetsci, viruses, water
%---------
% article
%---------
% The default type of manuscript is article, but can be replaced by: 
% addendum, article, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, correction, conferenceproceedings, conferencereport, expressionofconcern, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimage, letter, newbookreceived, opinion, obituary, projectreport, reply, retraction, review, preprints, shortnote, supfile, technicalnote
% supfile = supplementary materials
%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g. the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figure are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother 
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2018}
\copyrightyear{2018}
\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces}

% Authors, for the paper (add full first names)
\Author{Rodrigo Ramele $^{1,\dagger}$, Ana Julia Villar $^{1}$ and Juan Miguel Santos  $^{1}$*}
% Authors, for metadata in PDF
\AuthorNames{Rodrigo Ramele, Ana Julia Villar and Juan Miguel Santos}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
$^{1}$ \quad Computer Engineering Department, Instituto Tecnológico de Buenos Aires (ITBA); info@itba.edu.ar}

% Contact information of the corresponding author
\corres{Correspondence: rramele@itba.edu.ar; Tel.: +54-9-11-4193-9382}

% Current address and/or shared authorship
\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina}

@@

% Abstract (Do not use inserted blank lines, i.e. \\) 
\abstract{The Electroencephalography (EEG) is not just a mere clinical tool anymore.  It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time.  It is now being used to translate or decode brain signals to diagnose disease or implement Brain Computer Interface (BCI) devices to transmit volitional information.  The automatic decoding is mainly implemented by using quantitative algorithms to detect the cloaked information buried in the signal.  However, clinical EEG is based intensively on waveforms and the structure of signal plots. Hence, the purpose of this work is to establish a bridge to fill this gap by reviewing and describing the procedures that have been used to detect patterns in the electroencephalographic waveforms, benchmarking them on a controlled pseudo-real dataset of a P300-Based BCI Speller and verifying their performance on a public dataset of a BCI Competition.}

@@

% Keywords
\keyword{electroencephalography; brain-computer interfaces; waveform;p300;SIFT;PE;MP;SHCC}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{-1} %% Remove this when starting to work on the template.

@@

Current society is demanding technology to provide the means to realize the utopia of social inclusion for people with disabilities~\citep{Wolpaw2002}.  Additionally, as societies are aging~\citep{Lutz2008} the incidence of neuromuscular atrophies, strokes and other invalidating diseases is increasing.  Concurrently, the digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices~\citep{Domingo2012}.  All this human computer interaction is based on muscular movement~\citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human motion.  A new form of human machine communication which directly connects the Central Nervous System (CNS) to a machine or computer device is currently being developed: Brain Machine Interfaces (BMI), Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).

At the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the Electroencephalogram (EEG)~\citep{Schomer2010}.

EEG sensors are wearable~\citep{Puce2017} non-invasive, portable and mobile~\citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution~\citep{Hartman2005}.  This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write emails, control wheelchairs or to assess alcohol consumption~\citep{Yuste2017,KevricJasmin2016,Tzimourta2018,Cruz2018}.

@@

The EEG signal is a highly complex multi-channel time-series.  It can be modeled as a linear stochastic process with great similarities to noise ~\citep{Thakor2004}.  It is measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.  Figure \ref{fig:sampleeeg} shows $5$ seconds of a sample 8-channel EEG signal.

@@

The device that captures these small variations in potential differences over the scalp is called the Electroencephalograph.  Electrodes are located in predetermined positions over the head, usually embedded in saline solutions to facilitate the electrophysiological interface and are connected to a differential amplifier with a high gain which allows the measurement of tiny signals. Although initially analog devices were developed and used, nowadays digital versions connected directly to a computer are pervasive.  A detailed explanation on the particularities and modeling of EEG can be obtained from \citep{Jackson2014}, and a description of its electrophysiological aspects from \citep{Haberman2012}.

Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The following elements regularly  characterize EEG signals:

\begin{itemize}
\item Artifacts:  These are signal sources which are not generated from the CNS, but can be detected from the EEG signal.  They are called endogeneous or physiological when they are generated from a biological source like face muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic noise\citep{Weeda2012}.
\item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e. its mean and variance, and any other higher-order moments are not time-invariant \citep{Jansen1991}.
\item DC drift and trending: in EEG jargon, which is derived from concepts of electrical amplifiers theory, Direct Current (DC) refers to very low frequency components of the EEG signal which varies around a common center, usually the zero value.  DC drift means that this center value drifts in time.  Although sometimes considered as a nuisance that needs to get rid of, it is known that very important cognitive phenomena like Slow Cortical Potentials or Slow Activity Transients in infants do affect the drift and can be used to understand some particular brain functioning \cite{Schomer2010}.
\item Basal EEG activity: the EEG is the compound summation of myriads of electrical sources from the CNS.  These sources generate a baseline EEG which shows continuous activity with a small or null relation with any concurrent cognitive activity or task.
\item Inter-subject and intra-subject variability: EEG can be affected by the person's behavior like sleep hygiene, caffeine intake, smoking habit or alcohol intake previously to the signal measuring procedure \citep{Farzan2017}.
\end{itemize}

@@

\begin{itemize}
\item Spontaneous: generally treated as noise or basal EEG.
\item Evoked: activity that can be detected synchronously after some specific amount of time after the onset of the stimulus.  This is usually referred as time-locked.  In contrast to the previous one, it is often called Induced activity.
\end{itemize}

@@

\begin{itemize}
\item Rhythmic: EEG activity consisting in waves of approximately constant frequency.  It is often abbreviated RA (regular rythmic activity). They are loosely classified by their frequencies, and their naming convention was derived from the original naming used by Hans Berger himself, and after Alpha Waves (10 Hz), it came Delta (4 Hz), Theta (4-7 Hz), Sigma (12-16 Hz), Beta (12-30 Hz) and Gamma (30-100 Hz).  
\item Arrhythmic: EEG activity in which no stable rhythms are present.  
\item Dysrhythmic: Rhythms and/or patterns of EEG activity that characteristically appear in patient groups and rarely seen in healthy subjects.
\end{itemize}

The number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: signal elements can be generalized, focal or lateralized, depending on in which channel (i.e. electrode) they are found.

@@

\subsection{EEG Waveform Characterization}

@@

A waveform can have a characteristic shape, a rising or falling phase, a pronounced plateau or it may be composed of ripples and wiggles. In order to describe them, they are characterized by its amplitude, the arch, whether they have (non)sinusoidal shape, by the presence of an oscillation or imitating a sawtooth (e.g. Motor Cortical Beta Oscillations).  The characterization by their sharpness is also common, particularly in Epilepsy, and they can also be identified by their resemblance to spikes (e.g. Spike-wave discharge).

@@

\begin{itemize}
\item Attenuation: Also called suppression or depression. Reduction of amplitude of EEG activity resulting from decreased voltage. When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".
\item Hypersynchrony: Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range. The term suggest an increase in the number of neural elements contributing to the rhythm, or in the synchronization of different neurons with the same discharge pattern~\citep{Buzsaki2012}.
\item Paroxysmal: Activity that emerges from background with a rapid onset, reaching frequently high voltage and ending with an abrupt return to lower voltage activity. 
\item Monomorphic: Activity appearing to be composed of one dominant waveform pattern.
\item Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.
\item Transient/Component: An isolated wave or pattern that is distinctly different from background activity. 
\end{itemize}

@@

%\subsection{EEG Patterns}
%
%%The clinical EEG patterns are identified by the shape of waveform amplitude plotted against time.
%
%Non-exhaustive list of transient events
%Waveform characterization is of quite importance in terms of Event Related Potentials.  
%Determination of transient events, and particularly amplitude of different subcomponents, latency or even phase, has proved very importan concequences in terms of the different cognitive approach.
%
%%inverse problem is mathematically intractable Voytek 2009
%
%Oscillatory activity can also have their differente or distinctive waveforms.  Slow oscillations, which are assosiated with REM, sawtooth-shaped.
%Sleep spindles can also be considered oscilations and they have a distintinc form assosiated with stage 2 dream.
%Visual Cortical alpha and rolandic central mu waves arch-like structure (similar to the greek letter mu).  Slope Ratio.  Trough voltage remains contstant while peak voltage fluctuates.   Steep slopes,   Amplitude asymmetry 
%ponto-geniculo-occipital
%(PGO) waves.
%Rhythms in neural activity are observed across various temporal and spatial scales and are often
%referred to as oscillations (see Glossary) [1]. Traditionally, neural oscillations have been
%clustered into canonical frequency bands, including delta (1–4 Hz), theta (4–8 Hz), alpha (8–
%12 Hz), beta (15–30 Hz), gamma (30–90 Hz), and high gamma (>50 Hz). These bands roughly
%correspond to frequency ranges commonly observed in human electroencephalography (EEG)
%studies. Although they have been observed for nearly a century, recent theories suggest that
%these oscillations play an active role in neural communication.  Hippocampal theta oscillations, for example, are among the best-studied rhythms in the local field potential (LFP); they have a stereotyped sawtooth shape. In sleep research sleep 2 stage the background of KComplex and sleep spindles is theta waves
%
%%by application to Brain Computer Interfaces or EEG diagnosis (BCI a review, BCI Vidal 1973)
%
%
%
%%(e.g. FIRDA - Frontal Intermittent Rhythmic Delta) and posteriorly in children e.g. OIRDA - Occipital Intermittent Rhythmic Delta).
%
%%alpha dissapears when alerting by any mechanism (thinking, calculating)
%
%a) Spike: a transient with a pointed peak and a duration from 20 to under 70 msec.
%b) Sharp wave: a transient with a pointed peak and duration of 70-200 msec.
%
%Some initial works on EEG explored the idea to extend human capacities analyzing EEG waveforms  (automatic detection of k complexes), (A Waveform Analyzer Applied to the Human EEG) where a feature from amplitude and frequency of its signal and its derivative in time-domain is used.  Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.  
%
%ACA VA EL RESTO DE LOS METODOS (fujimori, PAA, etc).  Uchida 1999

@@

Sleep Research has been studied in this way by performing Polysomnographic recordings (PSG)~\citep{Boostani2017,Rodenbeck2006}. The different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized guidelines~\citep{Dimitriadis2018}. Visual characterization includes the identification or classification of certain waveform components based on a subjective characterization (e.g. positive or negative peak polarity) or the location within the strip.  It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g. sleep K-Complex is well characterized based on rates between positive vs negative amplitude)~\citep{Uchida1999}.  Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves, sleep spindles, polysplindles, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).

Moreover, EEG data acquisition is a key procedure during the assessment of patients with focal Epilepsy for potential seizure surgery, where the source of the seizure activity must be reliably identified. The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.  The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave discharges~\citep{EEGIntro}.

Waveform characterization is the method in which analysis has been performed for Event Related Potentials (ERP).  These are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.  ERPs are regularly used to assess auditory response in infants.  They are extensively used and studied in Cognitive Neuroscience~\citep{Luck2005}.  ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.   One of the most studied ERP is the P300, discovered in 1965 by Sutton, Braren, Zubin and John.  This component is a positive deflection of a subject's EEG signal that arises when an unexpected and infrequent stimulus appears~\citep{Wolpaw2002}.  The P300 is widely utilized in BCI because it can be harnessed to implement a Speller application.  Hence, P300 ERPs are a target phenomena to study by automatic waveform recognition methods.

>>>>

%=================================================================
\documentclass[brainsci,article,accept,moreauthors,pdftex,10pt,a4paper]{mdpi}

@@

%
%--------------------
% Class Options:
%--------------------
% journal
%----------
% Choose between the following MDPI journals:
% actuators, admsci, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, atmosphere, atoms, axioms, batteries, behavsci, beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci, buildings, carbon, cancers, catalysts, cells, challenges, chemosensors, children, chromatography, climate, coatings, computation, computers, condensedmatter, cosmetics, cryptography, crystals, data, dentistry, designs, diagnostics, diseases, diversity, econometrics, economies, education, electronics, energies, entropy, environments, epigenomes, fermentation, fibers, fishes, fluids, foods, forests, futureinternet, galaxies, games, gels, genealogy, genes, geosciences, geriatrics, healthcare, horticulturae, humanities, hydrology, informatics, information, infrastructures, inorganics, insects, instruments, ijerph, ijfs, ijms, ijgi, ijtpp, inventions, jcdd, jcm, jdb, jfb, jfmk, jimaging, jof, jintelligence, jlpea, jmse, jpm, jrfm, jsan, land, languages, laws, life, literature, lubricants, machines, magnetochemistry, marinedrugs, materials, mathematics, mca, mti, medsci, medicines, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, molbank, molecules, mps, nanomaterials, ncrna, neonatalscreening, nutrients, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, plants, polymers, processes, proteomes, publications, recycling, religions, remotesensing, resources, risks, robotics, safety, sensors, separations, sexes, sinusitis, socsci, societies, soils, sports, standards, sustainability, symmetry, systems, technologies, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vetsci, viruses, water
%---------
% article
%---------
% The default type of manuscript is article, but can be replaced by: 
% addendum, article, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, correction, conferenceproceedings, conferencereport, expressionofconcern, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimage, letter, newbookreceived, opinion, obituary, projectreport, reply, retraction, review, preprints, shortnote, supfile, technicalnote
% supfile = supplementary materials
%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figure are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother 
\articlenumber{x}
\doinum{10.3390/------}
\pubvolume{xx}
\pubyear{2018}
\copyrightyear{2018}
%\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1
\usepackage{subfigure}
\makeatletter
\renewcommand{\@thesubfigure}{\normalsize(\textbf{\alph{subfigure}})}
\makeatother
\usepackage{upgreek}
%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{EEG Waveform Analysis of P300 ERP with Applications to Brain Computer Interfaces}
%%Please carefully check the accuracy of names and affiliations. Changes will not be possible after proofreading. 
% Authors, for the paper (add full first names)
\Author{\hl{Rodrigo Ramele} *, Ana Julia Villar  and Juan Miguel Santos $^{\dagger}$}
% Authors, for metadata in PDF
\AuthorNames{Rodrigo Ramele, Ana Julia Villar and Juan Miguel Santos}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{%
Computer Engineering Department, Instituto Tecnológico de Buenos Aires (ITBA), \hl{Buenos Aires}, Argentina;  jsantos@itba.edu.ar (J.M.S.); jvillar@itba.edu.ar (A.J.V.)}
%%please add post code
% Contact information of the corresponding author
\corres{Correspondence: rramele@itba.edu.ar}

% Current address and/or shared authorship
\firstnote{Current address: C1437FBH Lavarden 315, Ciudad Autónoma de Buenos Aires, Argentina.}

@@

% Abstract (Do not use inserted blank lines, i.e., \\) 
\abstract{The Electroencephalography (EEG) is not just a mere clinical tool anymore.  It has become the de-facto mobile, portable, non-invasive brain imaging sensor to harness brain information in real time.  It is now being used to translate or decode brain signals, to diagnose diseases or to implement Brain Computer Interface (BCI) devices.  The automatic decoding is mainly implemented by using quantitative algorithms to detect the cloaked information buried in the signal.  However, clinical EEG is based intensively on waveforms and the structure of signal plots. Hence, the purpose of this work is to establish a bridge to fill this gap by reviewing and describing the procedures that have been used to detect patterns in the electroencephalographic waveforms, benchmarking them on a controlled pseudo-real dataset of a P300-Based BCI Speller and verifying their performance on a public dataset of a BCI Competition.}

@@

% Keywords
\keyword{electroencephalography; brain-computer interfaces; waveform; p300; SIFT; PE; MP;~SHCC}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\setcounter{section}{-1} %% Remove this when starting to work on the template.

@@

Current society is demanding technology to provide the means to realize the utopia of social inclusion for people with disabilities~\citep{Wolpaw2002}.  Additionally, as societies are aging~\citep{Lutz2008} the incidence of neuromuscular atrophies, strokes and other invalidating diseases is increasing.  Concurrently, the~digital revolution and the pervasiveness of digital gadgets have modified the way people interact with the environment through these devices~\citep{Domingo2012}.  All this human computer interaction is based on muscular movement~\citep{Guger2017}, but these trends are pushing this boundary beyond the confines of the body and beyond the limitation of human motion.  A new form of human machine communication which directly connects the Central Nervous System (CNS) to a machine or computer device is currently being developed: Brain Machine Interfaces (BMI), Brain Computer Interfaces (BCI) or Brain-Neural Computer Interfaces (BNCI).

At the center of all this hype, we can find a hundredth year old technology, rock-solid as a diagnosis tool, which greatly benefited from the shrinkage of sensors, the increase in computer power and the widespread development of wireless protocols and advanced electronics: the~Electroencephalogram (EEG)~\citep{Schomer2010}.

EEG sensors are wearable~\citep{Puce2017} non-invasive, portable and mobile~\citep{DeVos2014}, with excellent temporal resolution, and acceptable spatial resolution~\citep{Hartman2005}.  This humble diagnosis device is been transformed into currently the best approach to detect, out-of-the lab in an ambulatory context, information from the Central Nervous System and to use that information to volitionally drive cars, steer drones, write~emails, control wheelchairs or to assess alcohol \hl{consumption}~\citep{Yuste2017,KevricJasmin2016,Tzimourta2018,Cruz2018}.%%there are missing some references, we can not find the references,  please modify, please check all the text.

@@

The EEG signal is a highly complex multi-channel time-series.  It can be modeled as a linear stochastic process with great similarities to noise ~\citep{Thakor2004}.  It is measured in microvolts, and those slightly variations are contaminated with heavy endogenous artifacts and exogenous spurious signals.  Figure~\ref{fig:sampleeeg} shows $5$ s of a sample 8-channel EEG signal.

@@

The device that captures these small variations in potential differences over the scalp is called the Electroencephalograph.  Electrodes are located in predetermined positions over the head, usually embedded in saline solutions to facilitate the electrophysiological interface and are connected to a differential amplifier with a high gain which allows the measurement of tiny signals. Although initially analog devices were developed and used, nowadays digital versions connected directly to a computer are pervasive.  A detailed explanation on the particularities and modeling of EEG can be obtained from~\citep{Jackson2014}, and a description of its electrophysiological aspects from \citep{Haberman2012}.

Overall, EEG signals can be described by their phase, amplitude,  frequency and \textit{waveform}.  The~following elements regularly  characterize EEG signals:

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
\item Artifacts:  These are signal sources which are not generated from the CNS, but can be detected from the EEG signal.  They are called endogeneous or physiological when they are generated from a biological source like face muscles, ocular movements, etc., and exogeneous or non-physiological when they have an external electromagnetic source like line induced currents or electromagnetic~noise~\citep{Weeda2012}.
\item Non-Stationarity: the statistical parameters that describe the EEG as a random process are not conserved through time, i.e., its mean and variance, and any other higher-order moments are not time-invariant \citep{Jansen1991}.
\item DC drift and trending: in EEG jargon, which is derived from concepts of electrical amplifiers theory, Direct Current (DC) refers to very low frequency components of the EEG signal which varies around a common center, usually the zero value.  DC drift means that this center value drifts in time.  Although sometimes considered as a nuisance that needs to get rid of, it is known that very important cognitive phenomena like Slow Cortical Potentials or Slow Activity Transients in infants do affect the drift and can be used to understand some particular brain functioning \cite{Schomer2010}.
\item Basal EEG activity: the EEG is the compound summation of myriads of electrical sources from the CNS.  These sources generate a baseline EEG which shows continuous activity with a small or null relation with any concurrent cognitive activity or task.
\item Inter-subject and intra-subject variability: EEG can be affected by the person's behavior like sleep hygiene, caffeine intake, smoking habit or alcohol intake previously to the signal measuring~procedure \citep{Farzan2017}.
\end{itemize}

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
\item Spontaneous: generally treated as noise or basal EEG.
\item Evoked: activity that can be detected synchronously after some specific amount of time after the onset of the stimulus.  This is usually referred as time-locked.  In contrast to the previous one, it is often called Induced activity.
\end{itemize}

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
\item Rhythmic: EEG activity consisting in waves of approximately constant frequency.  It is often abbreviated RA (regular rythmic activity). They are loosely classified by their frequencies, and~their naming convention was derived from the original naming used by Hans Berger himself, and after Alpha Waves (10 Hz), it came Delta (4 Hz), Theta (4--7 Hz), Sigma (12--16~Hz), Beta~(12--30~Hz) and Gamma (30--100 Hz).  
\item Arrhythmic: EEG activity in which no stable rhythms are present.  
\item Dysrhythmic: Rhythms and/or patterns of EEG activity that characteristically appear in patient groups and rarely seen in healthy subjects.
\end{itemize}

The number of electrodes and their positions over the scalp determines a \textbf{Spatial Structure}: signal~elements can be generalized, focal or lateralized, depending on in which channel (i.e., electrode) they are found.

@@

\subsection*{EEG Waveform Characterization}

@@

A waveform can have a characteristic shape, a rising or falling phase, a pronounced plateau or it may be composed of ripples and wiggles. In order to describe them, they are characterized by its amplitude, the arch, whether they have (non)sinusoidal shape, by the presence of an oscillation or imitating a sawtooth (e.g., Motor Cortical Beta Oscillations).  The characterization by their sharpness is also common, particularly in Epilepsy, and they can also be identified by their resemblance to spikes (e.g., Spike-wave discharge).

@@

\begin{itemize}[leftmargin=*,labelsep=5.8mm]
\item Attenuation: Also called suppression or depression. Reduction of amplitude of EEG activity resulting from decreased voltage. When activity is attenuated by stimulation, it is said to have been "blocked" or to show "blocking".
\item Hypersynchrony: Seen as an increase in voltage and regularity of rhythmic activity, or within the alpha, beta, or theta range. The term suggest an increase in the number of neural elements contributing to the rhythm, or in the synchronization of different neurons with the same discharge~pattern~\citep{Buzsaki2012}.
\item Paroxysmal: Activity that emerges from background with a rapid onset, reaching frequently high voltage and ending with an abrupt return to lower voltage activity. 
\item Monomorphic: Activity appearing to be composed of one dominant waveform pattern.
\item \textls[-15]{Polymorphic: Activity composed of multiple frequencies that combine to form a complex waveform.}
\item Transient/Component:~An isolated wave or pattern that is distinctly different from background~activity. 
\end{itemize}

@@

%\subsection{EEG Patterns}
%
%%The clinical EEG patterns are identified by the shape of waveform amplitude plotted against time.
%
%Non-exhaustive list of transient events
%Waveform characterization is of quite importance in terms of Event Related Potentials.  
%Determination of transient events, and particularly amplitude of different subcomponents, latency or even phase, has proved very importan concequences in terms of the different cognitive approach.
%
%%inverse problem is mathematically intractable Voytek 2009
%
%Oscillatory activity can also have their differente or distinctive waveforms.  Slow oscillations, which are assosiated with REM, sawtooth-shaped.
%Sleep spindles can also be considered oscilations and they have a distintinc form assosiated with stage 2 dream.
%Visual Cortical alpha and rolandic central mu waves arch-like structure (similar to the greek letter mu).  Slope Ratio.  Trough voltage remains contstant while peak voltage fluctuates.   Steep slopes,   Amplitude asymmetry 
%ponto-geniculo-occipital
%(PGO) waves.
%Rhythms in neural activity are observed across various temporal and spatial scales and are often
%referred to as oscillations (see Glossary) [1]. Traditionally, neural oscillations have been
%clustered into canonical frequency bands, including delta (1–4 Hz), theta (4–8 Hz), alpha (8–
%12 Hz), beta (15–30 Hz), gamma (30–90 Hz), and high gamma (>50 Hz). These bands roughly
%correspond to frequency ranges commonly observed in human electroencephalography (EEG)
%studies. Although they have been observed for nearly a century, recent theories suggest that
%these oscillations play an active role in neural communication.  Hippocampal theta oscillations, for example, are among the best-studied rhythms in the local field potential (LFP); they have a stereotyped sawtooth shape. In sleep research sleep 2 stage the background of KComplex and sleep spindles is theta waves
%
%%by application to Brain Computer Interfaces or EEG diagnosis (BCI a review, BCI Vidal 1973)
%
%
%
%%(e.g., FIRDA - Frontal Intermittent Rhythmic Delta) and posteriorly in children e.g., OIRDA - Occipital Intermittent Rhythmic Delta).
%
%%alpha dissapears when alerting by any mechanism (thinking, calculating)
%
%a) Spike: a transient with a pointed peak and a duration from 20 to under 70 msec.
%b) Sharp wave: a transient with a pointed peak and duration of 70-200 msec.
%
%Some initial works on EEG explored the idea to extend human capacities analyzing EEG waveforms  (automatic detection of k complexes), (A Waveform Analyzer Applied to the Human EEG) where a feature from amplitude and frequency of its signal and its derivative in time-domain is used.  Althought CASENET REFERENCe explored "waveform" structure they were purely based on spike detection based on feeding artificial neural networks.  
%
%ACA VA EL RESTO DE LOS METODOS (fujimori, PAA, etc).  Uchida 1999

@@

Sleep Research has been studied in this way by performing Polysomnographic recordings (\hl{PSG})~\citep{Boostani2017,Rodenbeck2006}. The different sleep stages are evaluated by visually marking waveforms or graphoelements in long-running electroencephalographic recordings, looking for patterns based on standardized \hl{guidelines}~\citep{Dimitriadis2018}. Visual characterization includes the identification or classification of certain waveform components based on a subjective characterization (e.g., positive or negative peak polarity) or the location within the strip.  It is regular to establish an amplitude difference between different waveforms from which a relation between them is reckoned and a structured index is created (e.g., sleep K-Complex is well characterized based on rates between positive vs negative amplitude)~\citep{Uchida1999}.  Other relevant EEG patterns for sleep stage scoring are alpha, theta, and delta waves, sleep spindles, polysplindles, vertex sharp waves (VSW), and sawtooth waves (REM Sleep).

Moreover, EEG data acquisition is a key procedure during the assessment of patients with focal Epilepsy for potential seizure surgery, where the source of the seizure activity must be reliably identified. The onset of the Epileptic Seizure is defined as the first electrical change seen in the EEG rhythm which can be visually identified from the context and it is verified against any clinical sign indicating seizure onset.  The Interictal Epileptiform Discharges (IEDs) are visually identified from the paper strip, and they are also named according to their shape: spike, spike and wave or sharp-wave~discharges~\citep{EEGIntro}.

Waveform characterization is the method in which analysis has been performed for Event Related Potentials (ERP).  These are transient signal elements that may arise as a brain response to an external visual, tactile or auditory stimulus.  ERPs are regularly used to assess auditory response in infants.  They are extensively used and studied in Cognitive Neuroscience~\citep{Luck2005}.  ERPs are identified by their components which are recognizable signal shapes assigned to the observed waveform, that can be linked to some cognitive or measurable psychological process.   One of the most studied ERP is the P300, discovered in 1965 by Sutton, Braren, Zubin and John.  This component is a positive deflection of a subject's EEG signal that arises when an unexpected and infrequent stimulus appears~\citep{Wolpaw2002}.  The~P300 is widely utilized in BCI because it can be harnessed to implement a Speller application.  Hence, P300~ERPs are a target phenomena to study by automatic waveform recognition methods.

Table~\ref{tab:methods} summarizes a list of depictions used to describe waveforms in the surveyed literature.  Epilepsy has been described by the nature of oscillatory characterization of their waves, like ripples and wiggles, imitating sawtooths or by their geometric shape.  For ERPs on the other hand, more~elaborate indexes has been provided, establishing relations between amplitudes of signal components.  Finally, Sleep studies and ICU research are areas where the most complex indexes has been derived, particularly the coupling of signal properties like phase, amplitude and frequency.
